---
title: "Surgery Group 2"
name: "Ale, Daniella, Aaliyah"
date: "2026-02-27"
output:
  html_document:
    theme: spacelab
    toc: no

---

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(shiny)
library(shinythemes)
library(janitor)
library(scales)
```

## R Markdown

## Project Overview
This project analyzes whether surgery timing is associated with patient outcomes using data from 32,001 elective general surgery patients treated at Cleveland Clinic between 2005 and 2010.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

Surgery Timing: A dataset from 32,001 elective general surgery patients who were treated at Cleveland Clinic between 2005 and 2010. It is well known that lack of sleep and heavy workloads can negatively affect performance. Hospital staff frequently face these challenges, especially during busier periods. Because of this, researchers wanted to determine whether the time at which surgery is performed has any impact on patient outcomes.


```{r}
surgery <- read_csv("data/surgery.csv") %>% 
  clean_names()
```
```{r}
names(surgery)
```

1. Which race had the highest number of surgeries?
#Caucasians had the highest number of surgeries
```{r}
surgery %>% 
  group_by(race) %>% 
  summarise(
    n = n()
  ) %>% 
  arrange(desc(n))
```
```{r}
surgery %>% 
  count(race) %>% 
  ggplot(aes(x=race, y=n))+
  geom_col(fill= "turquoise4", color ="black", alpha = 0.9) +
  labs(title = "Number of Surgeries by Race",
       x = "Race",
       y = "Number of Surgeries")+
  theme_minimal()
```

2.At what age do people undergo surgery more? 
#90 year olds underwent surgery the most
```{r}
surgery %>% 
  group_by(age) %>% 
  summarise(
    n = n()
  ) %>% 
  arrange(desc(n))
```
```{r}
surgery %>% 
  ggplot(aes(x=age))+
  geom_histogram(fill= "turquoise4", color ="black", alpha = 0.9)+
  labs(title= "Distribution of Surgery by Age",
       x=NULL,
       y="# of Surgery's")
```

3.Who gets more surgeries, males or females ?
#Females get more surgeries than males
```{r}
surgery %>% 
  filter(gender != "NA") %>% 
  group_by(gender) %>% 
  summarise(
    n = n()
  ) %>% 
  arrange(desc(n))
```

```{r}
 surgery %>% 
  filter(gender != "NA") %>% 
  count(gender) %>% 
  ggplot(aes(x = gender, y = n, fill = gender)) +
  geom_col(fill= "turquoise4", color ="black", alpha = 0.9) +
  labs(title = "Number of Males and Females",
       x = "Gender",
       y = "Count")+
  theme_minimal()
```

4.Is mortality higher in males or females?
#Mortality is higher in males than in females
```{r}
surgery %>%
  filter(gender != "NA") %>%
  count(gender, mort30)
```

```{r}
surgery %>% 
  filter(gender != "NA") %>% 
  group_by(gender) %>% 
  summarize(deaths = sum(mort30 == "Yes", na.rm = TRUE)) %>% 
  ggplot(aes(gender, deaths)) + 
  geom_col(fill="turquoise4", color="black") + 
  labs( title="Deaths by Gender", x="Gender", y="Number of Deaths" ) + 
  theme_minimal()
```


5.Does ASA status predict mortality?
#ASA statuses IV-VI have the highest mortality rates, albeit it is only 4%
```{r}
surgery %>%
  filter(asa_status!="NA") %>% 
  group_by(asa_status) %>%
  summarise(mortality_rate = mean(mort30 == "Yes", na.rm = T))
```

```{r}
surgery %>%
  filter(asa_status!="NA") %>%
  group_by(asa_status) %>%
  summarise(mortality_rate=mean(mort30=="Yes", na.rm=TRUE)) %>%
  ggplot(aes(asa_status, mortality_rate))+
  geom_col(fill="turquoise4", color="black", alpha=0.9)+
  scale_y_continuous(labels=scales::percent)+
  labs(title="Mortality Rate by ASA Status",
       x="ASA Status",
       y="Mortality Rate")+
  theme_minimal()
```


6. What is the distribution of age for each ASA status? Is there a relationship?
#The older someone seems to get, the higher the ASA status is, with the density graphs skewing further left the higher the ASA status levels go
```{r}
surgery %>%
  filter(asa_status %in% c("I-II", "III", "IV-VI")) %>%
  ggplot(aes(x = age)) +
  geom_density(color = "turquoise4", fill = "turquoise4", alpha = 0.4) +
  facet_wrap(~ asa_status) +
  labs(
    title = "Distribution of Age by ASA Status",
    x = "Age",
    y = "Density"
  ) 
```


7.Do 30-day mortality rates differ by type of surgical procedure?
```{r}
surgery %>% 
  group_by(ahrq_ccs) %>% 
  summarise(ccsmort30rate = mean(ccsmort30rate, na.rm = TRUE)) %>% 
  ggplot(aes(x = reorder(ahrq_ccs, ccsmort30rate), y = ccsmort30rate)) +
  geom_col(fill = "turquoise4") +
  coord_flip() +
  labs(title = "30-Day Mortality by Procedure",
       x = NULL,
       y = "Rate") +
  theme_light()
```

8.What month is mortality highest in and what month is the best for surgery?
#January appears to have the highest mortality rate, while November appears to have the lowest mortality rate, meaning november is the best month to have surgery. 
```{r}
surgery %>%
  filter(month != "NA") %>%
   mutate(month = factor(month, c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"))) %>%
  group_by(month) %>%
  summarize(mortality = sum(mort30 == "Yes", na.rm = T)) %>%
  arrange(month)
```


```{r}
surgery %>%
  filter(month!="NA") %>%
  mutate(month=factor(month,
                      c("Jan","Feb","Mar","Apr","May","Jun",
                        "Jul","Aug","Sep","Oct","Nov","Dec"))) %>%
  group_by(month) %>%
  summarise(mortality_rate=mean(mort30=="Yes", na.rm=TRUE)) %>%
  ggplot(aes(month, mortality_rate, group=1))+
  geom_line(color="turquoise4", linewidth=1.2)+
  geom_point(size=3, color="turquoise4")+
  scale_y_continuous(labels=scales::percent)+
  labs(title="Mortality Rate by Month",
       x="Month",
       y="Mortality Rate")+
  theme_minimal()
```
9.Is there any association between moon phase and mortality?
#Full moons have the highest mortality rate, while the new moon has the lowest mortality rate
```{r}
moon <- surgery %>% 
  group_by(moonphase) %>% 
  summarize(
    mortality_rate = mean(mort30 == "Yes", na.rm = T),
    n = n(),
    .groups = "drop")
```

```{r}
moon %>% 
  ggplot(aes(x = moonphase,
             y = mortality_rate,
             group = 1)) +
  geom_point(size = 3, color = "turquoise4") +
  geom_line(color = "turquoise4", linewidth = 1) +
  theme_minimal() + 
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Mortality Rate by Moonphase",
    x = NULL,
    y = "Mortality Rate"
  )
```

10. What surgery hour is associated with the highest mortality rate?
#The mortality rate varies slightly depending on the hour of surgery. Surgeries performed earlier in the day tend to have lower mortality rates, while surgeries later in the afternoon show slightly higher mortality rates, suggesting that earlier surgery times may be associated with better patient outcomes.
```{r}
surgery %>% 
  mutate(hour_round = floor(hour)) %>% 
  group_by(hour_round) %>% 
  summarize(mortality_rate = mean(mort30 == "Yes", na.rm = TRUE)) %>% 
  ggplot(aes(x = factor(hour_round), y = mortality_rate)) +
  geom_col(color = "black", fill = "turquoise4", alpha = 0.8) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Mortality Rate by Operation Hour",
    x = "Hour (Military Time)",
    y = "Mortality Rate"
  ) +
  theme_minimal()
```

11.Does the day of the week that a surgery is performed affect patient mortality?
```{r}
surgery %>%
  filter(dow != "NA") %>%
  mutate(dow = factor(dow, levels = c("Mon","Tue","Wed","Thu","Fri"))) %>%
  group_by(dow) %>%
  summarise(mortality_rate = mean(mort30 == "Yes", na.rm = TRUE)) %>%
  ggplot(aes(x = dow, y = mortality_rate)) +
  geom_col(fill = "turquoise4", color = "black", alpha = 0.9) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Mortality Rate by Day of Week",
    x = "Day of Week",
    y = "Mortality Rate"
  ) +
  theme_minimal()
```

12.Shiny App

```{r, eval=F, echo=T}
ui <- fluidPage(
  titlePanel("Surgery Timing and Patient Outcomes"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput(
        "timing",
        "Choose Timing Variable:",
        choices = c("Hour","Day of Week","Month","Moon Phase"),
        selected = "Hour"
      )
    ),
    
    mainPanel(
      plotOutput("plot", width="700px", height="500px")
    )
  )
)

server <- function(input, output, session) {
  
  output$plot <- renderPlot({
    
    if(input$timing == "Hour"){
      
      surgery %>% 
        mutate(hour_round = floor(hour)) %>% 
        group_by(hour_round) %>% 
        summarize(mortality_rate = mean(mort30 == "Yes", na.rm = TRUE)) %>% 
        ggplot(aes(x = factor(hour_round), y = mortality_rate))+
        geom_col(color="black", fill="turquoise4", alpha=0.8)+
        scale_y_continuous(labels=scales::percent)+
        labs(title="Mortality Rate by Operation Hour",
             x="Hour",
             y="Mortality Rate")+
        theme_minimal()
      
    } else if(input$timing == "Month"){
      
      surgery %>%
        filter(month!="NA") %>%
        mutate(month=factor(month,
                            levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                     "Jul","Aug","Sep","Oct","Nov","Dec"))) %>%
        group_by(month) %>%
        summarise(mortality_rate=mean(mort30=="Yes", na.rm=TRUE)) %>%
        ggplot(aes(month, mortality_rate, group=1))+
        geom_line(color="turquoise4", linewidth=1.2)+
        geom_point(size=3, color="turquoise4")+
        scale_y_continuous(labels=scales::percent)+
        labs(title="Mortality Rate by Month",
             x="Month",
             y="Mortality Rate")+
        theme_minimal()
      
    } else if(input$timing == "Day of Week"){
      
      surgery %>%
        mutate(dow=factor(dow,
                          levels=c("Mon","Tue","Wed","Thu","Fri"))) %>%
        group_by(dow) %>%
        summarize(mortality_rate=mean(mort30=="Yes", na.rm=TRUE)) %>%
        ggplot(aes(dow, mortality_rate))+
        geom_col(fill="turquoise4", color="black")+
        scale_y_continuous(labels=scales::percent)+
        labs(title="Mortality Rate by Day of Week",
             x="Day",
             y="Mortality Rate")+
        theme_minimal()
      
    } else {
      
      surgery %>% 
        group_by(moonphase) %>% 
        summarize(mortality_rate = mean(mort30=="Yes", na.rm=TRUE)) %>%
        ggplot(aes(moonphase, mortality_rate, group=1))+
        geom_point(size=3, color="turquoise4")+
        geom_line(color="turquoise4")+
        scale_y_continuous(labels=scales::percent)+
        labs(title="Mortality Rate by Moon Phase",
             x="Moon Phase",
             y="Mortality Rate")+
        theme_minimal()
      
    }
    
  })
}

shinyApp(ui, server)
```

```{r}
library(shiny)

ui <- fluidPage(
  theme = shinytheme("flatly"),
  
  tags$head(
    tags$style(HTML("body { background-color: #f8f9fa;}
    .app-title { 
    text-align:center;
    font-size: 34px;
    font-weight: 700;
    color: #d63384;
    margin-top: 15px;
    text-align:center;
    box-shadow: 2px 2px 8px rgba (0,0,0,0.08);
    }
    .stat-label {
    font-size:15px;
    color: #6c757d;
    }
    .stat-value {
    font-size: 24px;
    font-weight: bold;
    color: #d63384;
    }
    .sidebar-panel {
    background-color: #ffffff;
    padding:15px;
    border-radius: 12px;
    box-shadow: 2px 2px 8px rgba(0,0,0,0.08)
    }
    .main-panel-box{
    background-color: #ffffff;
    padding: 15px;
    border-radius: 12px;
    box-shadow: 2px 2px 8px rgba(0,0,0,0.08)
    }
    "))),
  
  div(class = "app-title", "Surgery Timing and Patient Outcomes"),
  
  fluidRow(
    column(4,
           div(class = "stat-box",
               div(class = "stat-label", "Highest Mortality Hour"),
               div(class = "stat-value", textOutput("highest_hour")))),
    column(4,
           div(class = "stat-box",
               div(class = "stat-label", "Lowest Mortality Hour"),
               div(class = "stat-value", textOutput("lowest_hour")))),
     column(4,
           div(class = "stat-box",
               div(class = "stat-label", "Overall Mortality Rate"),
               div(class = "stat-value", textOutput("overall_mortality"))))
  ),
    
    sidebarLayout(
      sidebarPanel(
        div(class = "sidebar-panel",
        h4("Plot Controls"),
        
        selectInput(
          "timing",
          "Choose Timing Variable",
          choices = c("Hour", "Day of Week", "Month", "Moonphase"),
          selected = "Hour"
        ),
        
        selectInput(
          "plot_color",
          "Choose Plot Color",
          choices = c("Pink" = "hotpink",
                      "Blue" = "steelblue",
                      "Green" = "darkgreen",
                      "Purple" = "purple",
                      "Orange" = "orange"),
          selected = "steelblue"
        ),
        
        checkboxInput("show_points", "Show Points on Graph", value = T)
      ),
      
      width = 3
    ),
    
    mainPanel(
      div(class = "main-panel-box",
          tabsetPanel(
            tabPanel(
              "Plot",
              br(),
              plotOutput("plot", width = "100%", height = "520px")
            ),
            tabPanel(
              "About",
              br(),
              h4("About Our App"),
              p("This app explores the relationship between different time variables during surgery and patient mortality."),
              tags$ul(
                tags$li("Hour"),
                tags$li("Day of Week"),
                tags$li("Month"),
                tags$li("Moonphase")
              ),
              p("Use the sidebar to toggle variables and colors")
            )
          )
      ),
      width = 9 
    )
  )
)

server <- function(input, output, session) {
  
  hour_data <- reactive({
    surgery %>% 
      mutate(hour_round = floor(hour)) %>% 
      group_by(hour_round) %>% 
      summarize(mortality_rate = mean(mort30 == "Yes", na.rm = T), .groups = "drop")
  })
  
  output$highest_hour <- renderText({
    highest <- hour_data() %>% 
      filter(mortality_rate == max(mortality_rate, na.rm = T))
    
    paste0(highest$hour_round[1], ":00")
  })
  
   output$lowest_hour <- renderText({
    lowest <- hour_data() %>% 
      filter(mortality_rate == min(mortality_rate, na.rm = T))
    
    paste0(lowest$hour_round[1], ":00")
   })
   
    output$overall_mortality <- renderText({
    rate <- mean(surgery$mort30 == "Yes", na.rm = T) 
      scales::percent(rate, accuracy = 0.01)
    })
    
    output$plot <- renderPlot({
      
      if (is.null(input$timing) || input$timing == "") {
        return(NULL)
      }
      
      if (input$timing == "Hour") {
        plot_data <- surgery %>% 
          mutate(hour_round = floor(hour)) %>% 
          group_by(hour_round) %>% 
          summarize(mortality_rate = mean(mort30 == "Yes", na.rm = T), .groups = "drop")
        
        ggplot(plot_data, aes(x = factor(hour_round), y = mortality_rate))+
          geom_col(fill = input$plot_color, color = "black", alpha = 0.9)+
          scale_y_continuous(labels = percent)+
          labs( title = "Mortality Rate by Operation Hour",
                x = "Hour",
                y = "Mortality Rate")+
          theme_minimal()+
          theme(
            plot.title = element_text(face = "bold", size = 18, hjust = 0.5, color = "maroon"),
            axis.title = element_text(face = "bold"),
            panel.grid.minor = element_blank()
          )
      }
      
      else if (input$timing == "Day of Week")
      {
        plot_data <- surgery %>% 
          mutate(dow = factor(dow, levels = c("Mon", "Tue", "Wed", "Thu", "Fri"))) %>% 
          group_by(dow) %>% 
          summarize(mortality_rate = mean(mort30 == "Yes", na.rm = T), .groups = "drop")
        
         ggplot(plot_data, aes(x = dow, y = mortality_rate))+
          geom_col(fill = input$plot_color, color = "black", alpha = 0.9)+
          scale_y_continuous(labels = percent)+
          labs( title = "Mortality Rate by DOW",
                x = "Day of Week",
                y = "Mortality Rate")+
          theme_minimal()+
          theme(
            plot.title = element_text(face = "bold", size = 18, hjust = 0.5, color = "maroon"),
            axis.title = element_text(face = "bold"),
            panel.grid.minor = element_blank()
          )
         }
      else if (input$timing == "Month")
      {
        plot_data <- surgery %>% 
          filter(month != "NA") %>% 
          mutate(month = factor(month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "June",
                                          "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))) %>% 
          group_by(month) %>% 
          summarize(mortality_rate = mean(mort30 == "Yes", na.rm = T), .groups = "drop")
        
         p <- ggplot(plot_data, aes(x = month, y = mortality_rate, group = 1))+
          geom_line(color = input$plot_color)
         
         if (input$show_points){
           p <- p + geom_point (size = 4, color = input$plot_color)
         }
         
         p+
          scale_y_continuous(labels = percent)+
          labs( title = "Mortality Rate by Month",
                x = "Month",
                y = "Mortality Rate")+
          theme_minimal()+
          theme(
            plot.title = element_text(face = "bold", size = 18, hjust = 0.5, color = "maroon"),
            axis.title = element_text(face = "bold"),
            panel.grid.minor = element_blank()
          )
    }
         else {
           plot_data <- surgery %>% 
             group_by(moonphase) %>% 
              summarize(mortality_rate = mean(mort30 == "Yes", na.rm = T), .groups = "drop")
           
            p <- ggplot(plot_data, aes(x = moonphase, y = mortality_rate, group = 1))+
          geom_line(color = input$plot_color)
         
         if (isTRUE(input$show_points)){
           p <- p + geom_point (size = 4, color = input$plot_color)
         }
           
           p +
          scale_y_continuous(labels = percent)+
          labs( title = "Mortality Rate by Moonphase",
                x = "Phase of Moon",
                y = "Mortality Rate")+
          theme_minimal()+
          theme(
            plot.title = element_text(face = "bold", size = 18, hjust = 0.5, color = "maroon"),
            axis.title = element_text(face = "bold"),
            panel.grid.minor = element_blank()
          )
         }
    })
}

shinyApp(ui, server)
```


